Dark Control: The Default Mode Network as a Reinforcement Learning Agent

### 2018-08-11 paper/paper.tex
Update Section 3.3 & VennDiagram

\section{Introduction}
%
In the absence of external stimulation, the human brain is not at rest.
At the turn to the 21st century,
brain-imaging may have been the first technique
to allow for the discovery of a unique brain network that would subserve
baseline mental activities
\citep{raichle2001pnas, randy2008, bzdok2015resting}.
The ``default mode network'' (DMN) continues
to metabolize large quantities of
oxygen and glucose energy to maintain
neuronal computation
during free-ranging thought
\citep{kenet2003spontaneously, fiser2004small}.
The baseline energy demand is only weakly modulated
at the onset of defined psychological tasks \citep{raichle_baseline}.
At its opposite, during sleep, the
decoupling of brain structures discarded the idea of the DMN being
only a passive network resonance and rather supported an
important role in sustaining conscious awareness \citep{horovitz2009decoupling}.

@@

\subsection{Towards a formal account of default mode function: higher-order control of the organism}
The network nodes that compose the human DMN
are hubs of high baseline neural activity,
which typically decreases when engaged in controlled psychological experiments
\citep{raichle_baseline}.
The standard mode of
neural information maintenance and manipulation has
been argued to mediate evolutionarily conserved functions
\citep{brown1914nature, binder1999conceptual, buzsaki2006rhythms}.
Today, many psychologists and neuroscientists believe that
the DMN implements some form of
probabilistic estimation of past, hypothetical, and
future events
\citep{fox2005, hassabis2007patients, schacter2007remembering, binder2009,
randy2008, spreng2009common}.
This brain network
might have emerged to continuously predict the environment using
mental imagery as an evolutionary advantage \citep{suddendorf2007evolution}.
%
However, information processing in the DMN has also repeatedly
been shown to directly impact human behavior. Goal-directed task performance
improved with decreased activity in default mode regions \citep{weiss2006}
and increased DMN activity was linked to more task-independent,
yet sometimes useful thoughts
\citep{mason2007, seli2016mind}.
%
Gaining insight into DMN function is
particularly challenging because
this brain network appears to
simultaneously modulate perception-action cycles in the present
and to support mental travel
across time, space, and content domains
\citep{boyer2008evolutionary}.

@@

\subsubsection{Accumulated rewards and policies}
The behavior of the agent is governed by a \textit{policy}, which maps states of the world
to probability distributions over candidate actions.
Starting at time $t=0$,
following a policy $\pi$ generates a trajectory of action choices:
\begin{eqnarray*}
  \begin{split}
    \textbf{choose action: }a_0 &\sim \pi(a|s_0)\\
    \textbf{ observe transition: }s_1 &\sim p(s|s_0,a_0)\textbf{ and collect
      reward }R_0 = r(s_0, a_0, s_1)\\
    \textbf{choose action: }a_1 &\sim \pi(a|s_1)\\
    \textbf{ observe transition: }s_2 &\sim p(s|s_1,a_1), \textbf{ and collect
      reward }R_1 = r(s_1, a_1,s_2)\\
    \vdots\\
    \textbf{choose action: }a_{t} &\sim \pi(a|s_{t})\\
    \textbf{ observe transition: }s_{t+1} &\sim p(s|s_{t},a_{t}), \textbf{ and
      collect reward }R_{t} = r(s_{t}, a_{t},s_{t+1})\\
    \vdots
  \end{split}
\end{eqnarray*}
We assume time invariance
in that we expect the dynamics of the process
to be equivalent over sufficiently long time windows of equal length (i.e., stationarity).
Since an action executed in the present moment might have repercussions in the
far future, it turns out that the quantity to optimize is not the instantaneous
rewards $r(s, a)$, but a \textit{cumulative reward} estimate which takes into
account expected reward from action choices in the future. A common approach to
modeling this gathered outcome is the time-discounted cumulative reward% function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t =
  R_0 + \gamma R_1 + \gamma^2 R_2 + \ldots + \gamma^tR_t + \ldots
\end{equation}
This random variable
  \footnote{The reward outcome is random as it depends both on the environment's dynamics and the
  policy $\pi$ being executed. The exponential delay discounting function used
  here refers to the usual formulation in the field of reinforcement learning,
  although psychological experiments may also reveal other discounting regimes
  \citep{green2004discounting}.}
measures the cumulative reward of
following an action policy $\pi$.
%
Note that value buffering may be realized in the vmPFC by virtue of
this region's connections to the NAc of the reward system
~\citep{carlezon2009,haber1995,croxson2005quantitative}.

@@

\subsubsection{Optimal control via stochastic gradient descent in the DMN}
Efficient learning of the entire set of model parameters can effectively be achieved
via stochastic \textit{gradient descent}, a universal algorithm for finding
local minima based on the first derivative of the optimization objective.
Stochastic here means that the gradient is estimated from batches of training samples,
which here corresponds to blocks of experience from the replay memory:

@@

\subsection{Putting everything together: hypotheses for future studies}

@@

\paragraph*{Neural correlates of the Bellman equation in the DMN (Humans)}
Relating decomposition of consecutive action choices by the Bellman equation
to neuroscientific insights,
specific neural activity in the dorsal prefrontal cortex (BA9)
was linked to processing ``goal-tree sequences''
in human brain-imaging experiments
\citep{koechlin1999role, koechlin2000dissociating}.
Sub-goal exploration may require multi-task switching
between cognitive processes
as later parts of a solution
frequently depend on respective earlier steps in a given solution path, which
necessitates storage of expected intermediate outcomes.
As such,
``cognitive branching'' operations for nested processing of behavioral strategies
are likely to
entail secondary reallocation of attention and working-memory resources.
Further brain-imaging experiments corroborated
the prefrontal DMN to subserve
``processes related to the management and monitoring of sub-goals while
maintaining information in working memory''
\citep{braver2002role}
and to functionally couple with the hippocampus conditioned by ``deep versus shallow planning''
\citep{kaplan2017neural}.
%
Moreover,
neurological patients with lesions in this DMN region were reported
to be impaired in aspects of realizing ``multiple sub-goal scheduling''
\citep{burgess2000cognitive}.
Hence,
the various advanced human abilities subserved by the DMN, such as
planning and abstract reasoning, can be viewed to involve some form of
action-decision branching to enable higher-order executive control.

We hypothesize a functional dissociation between computations pertaining to action policy versus adapting stimulus-value associations as we expect implementation in different subsystems of the DMN. First, we expect that fMRI signals in the right temporo-parietal junction relate to behavioral changes subsequent to adaptation in the action choice tendencies (policy matrix) involved in non-value-related prediction error. Second, fMRI signals in the ventromedial prefrontal cortex should relate to behavioral changes following adaptation in value estimation (value matrix) due to reward-related stimulus-value association. We finally expect that fMRI signals in the posteromedial cortex, as a potential global information integrator, are related to shifts in overt behavior based on previous adaptations in both policy or value estimation.

\paragraph*{Continuous learning paradigm (Humans)} We hypothesize a functional relationship between the DMN closely associated with the occurrence of stimulus-independent thoughts and the reward circuitry. During an iterative neuroeconomic two-player game, fMRI signals in the DMN could be used to predict reward-related signals in the nucleus accumbens across trials in a continuous learning paradigm. We expect that the more DMN activity is measured to be increased, supposedly the higher the tendency for stimulus-independent thoughts, the more the fMRI signals in the reward circuits should be independent of the reward context in the current sensory environment.

\paragraph*{Temporal discounting experiment (Humans)} We hypothesize that the relevant time horizon is modulated by various factors such as age, acute stress and time-enduring impulsivity traits
\citep{luksys2009stress, haushofer2014psychology}.
Using a temporal discounting experiment, it can be quantified how the time horizon is affected at the behavioral level and then traced back to its corresponding neural representation. Such experimental investigation can be designed to examine between-group and within-group effects (e.g., impulsive population like chronic gamblers or drug addicts); and brought in context with the participant’s age and personality traits.
% To exam-ine this question, Fellows and Farah (2005) investigated the effects of dorsolateral frontal (DLF) and ventromedial frontal (VMF) lobe damage on two aspects of temporal foresight—namely, temporal discounting and the futuretime perspective (measuredby an adaptation of Wallace’s [1956] list of future life events).

@@

The HC may contribute to generating
perturbed action-transition-state-reward samples
as batches of pseudo-experience
(i.e., recalled, hypothesized, and forecasted scenarios).
The small variations in these experience samplings allow searching
a larger space of model parameters and possible experiences.
Taken to its extreme, stochastic recombination of experience
building blocks can further optimize the behavior of the RL agent
by learning from scenarios in the environment that the agent might
encounter only very rarely or never.
An explanation is thus offered for experiencing seemingly familiar situations that
a human has however never actually encountered (i.e., d\'{e}j\`{a} vu effect).
While such a situation may not have been experienced in the physical world,
the DMN may have previously stochastically generated, evaluated, and adapted to
such a randomly synthesized event.
Generated representations arguably are ``internally manipulable,
and can be used for attempting actions internally,
before or instead of acting in the external reality,
and in diverse goal and sensory contexts, i.e. even outside the context in which they were learned''
\citep{pezzulo2011grounding}.
In the context of scarce environmental input and feedback
(e.g., mind-wandering or sleep),
mental scene construction allows pseudo-experiencing possible
future scenarios and action outcomes.

\paragraph*{Interplay with memory and “mind-searching” (Animals)} We hypothesize that experience replay for browsing problem solutions subserved by the DMN is contributing to choice behavior in mice. Hippocampal single-cell recordings have shown that neural patterns during experimental choice behavior are reiterated during sleep and before making analogous choices in the future. Necessity of cortical DMN regions, in addition to the hippocampus, for “mind-searching” candidate actions during choice behavior can be experimentally corroborated by causal disruption of DMN regions, such as by circumscribed brain lesion or optogenetic intervention in the inferior parietal and prefrontal cortices.

From the perspective of a RL agent,
prediction in the DMN reduces to
generalization of
policy and value computations from sampled experiences to
successful action choices and reward predictions in future states.
As such,
plasticity in the DMN arises naturally.
If an agent behaving optimally in a certain environment moves
to new, yet unexperienced environment, reward prediction errors will
largely increase.
This feedback will lead to adaptation of policy considerations
and value estimations until the intelligent system converges to a
new steady state of optimal action decisions in a volatile world.

\paragraph*{Synaptic epigenesis and creation of new entry in policy matrix (Humans \& Animals)} An additional layer of learning concerns the addition of new entries in the state and action spaces. Extension of the action repertoire could be biologically realized by synaptic epigenesis \citep{gisiger_acquisition_2005}. Indeed, the tuning of synaptic weights through learning can stabilize additional patterns of activity by creating new attractors in the neural dynamics landscape \citep{takeuchi_synaptic_2014}. Those attractors can then constrain both the number of factors taken into account by decision processes and the possible behaviors of the agent \citep{wang_decision_2008}. To examine this potential higher-level mechanism, we propose to probe how synaptic epigenesis is related to neural correlates underlying policy matrix updates: in humans the changes of functional connectivity between DMN regions can be investigated following a temporal discounting experiment and in monkeys or rodents anterograde tracing can be used to study how homolog regions of the DMN present increased synaptic changes compare to other parts of the brain.

@@

\begin{figure}[!h]
  \includegraphics[width=1\linewidth]{VennDiagram-2017-09-08.pdf}
  \caption{\textbf{Situating Markov Decision Processes among other accounts of default mode function.}
  The Venn diagram summarizes the relationship between four previously proposed explanations for the functional role of the DMN and our present account. Viewing empirical findings in the DMN from the MDP viewpoint incorporates important aspects of the free energy principle, predictive coding, sentinal hypothesis, and semantic hypothesis.
  The MDP account may reconcile several strengths of these functional accounts in a process model that simultaneously acknowledges environmental input and behavioral choices as well as the computational and algorithmic properties (How? and What?) underlying higher-order control of the organism.}
  \label{fig:VennDiagram}
\end{figure}

>>>>

\section{Towards a formal account of default mode function: higher-order control of the organism}
%
In the absence of external stimulation, the human brain is not at rest.
At the turn to the 21st century,
brain-imaging may have been the first technique
to allow for the discovery of a unique brain network that would subserve
baseline mental activities
\citep{raichle2001pnas, randy2008, bzdok2015resting}.
The ``default mode network'' (DMN) continues
to metabolize large quantities of
oxygen and glucose energy to maintain
neuronal computation
during free-ranging thought
\citep{kenet2003spontaneously, fiser2004small}.
The baseline energy demand is only weakly modulated
at the onset of defined psychological tasks \citep{raichle_baseline}.
At its opposite, during sleep, the
decoupling of brain structures discarded the idea of the DMN being
only a passive network resonance and rather supported an
important role in sustaining conscious awareness \citep{horovitz2009decoupling}.

@@

% \subsection{Towards a formal account of default mode function: higher-order control of the organism}
The network nodes that compose the human DMN
are hubs of high baseline neural activity,
which typically decreases when engaged in controlled psychological experiments
\citep{raichle_baseline}.
The standard mode of
neural information maintenance and manipulation has
been argued to mediate evolutionarily conserved functions
\citep{brown1914nature, binder1999conceptual, buzsaki2006rhythms}.
Today, many psychologists and neuroscientists believe that
the DMN implements some form of
probabilistic estimation of past, hypothetical, and
future events
\citep{fox2005, hassabis2007patients, schacter2007remembering, binder2009,
randy2008, spreng2009common}.
This brain network
might have emerged to continuously predict the environment using
mental imagery as an evolutionary advantage \citep{suddendorf2007evolution}.
%
However, information processing in the DMN has also repeatedly
been shown to directly impact human behavior. Goal-directed task performance
improved with decreased activity in default mode regions \citep{weiss2006}
and increased DMN activity was linked to more task-independent,
yet sometimes useful thoughts
\citep{mason2007, seli2016mind}.
%
Gaining insight into DMN function is
particularly challenging because
this brain network appears to
simultaneously modulate perception-action cycles in the present
and to support mental travel
across time, space, and content domains
\citep{boyer2008evolutionary}.

@@

\paragraph*{Accumulated rewards and policies}
The behavior of the agent is governed by a \textit{policy}, which maps states of the world
to probability distributions over candidate actions.
Starting at time $t=0$,
following a policy $\pi$ generates a trajectory of action choices:
\begin{eqnarray*}
  \begin{split}
    \textbf{choose action: }a_0 &\sim \pi(a|s_0)\\
    \textbf{ observe transition: }s_1 &\sim p(s|s_0,a_0)\textbf{ and collect
      reward }R_0 = r(s_0, a_0, s_1)\\
    \textbf{choose action: }a_1 &\sim \pi(a|s_1)\\
    \textbf{ observe transition: }s_2 &\sim p(s|s_1,a_1), \textbf{ and collect
      reward }R_1 = r(s_1, a_1,s_2)\\
    \vdots\\
    \textbf{choose action: }a_{t} &\sim \pi(a|s_{t})\\
    \textbf{ observe transition: }s_{t+1} &\sim p(s|s_{t},a_{t}), \textbf{ and
      collect reward }R_{t} = r(s_{t}, a_{t},s_{t+1})\\
    \vdots
  \end{split}
\end{eqnarray*}
We assume time invariance
in that we expect the dynamics of the process
to be equivalent over sufficiently long time windows of equal length (i.e., stationarity).
Since an action executed in the present moment might have repercussions in the
far future, it turns out that the quantity to optimize is not the instantaneous
rewards $r(s, a)$, but a \textit{cumulative reward} estimate which takes into
account expected reward from action choices in the future. A common approach to
modeling this gathered outcome is the time-discounted cumulative reward% function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t =
  R_0 + \gamma R_1 + \gamma^2 R_2 + \ldots + \gamma^tR_t + \ldots
\end{equation}
This random variable
  \footnote{The reward outcome is random as it depends both on the environment's dynamics and the
  policy $\pi$ being executed. The exponential delay discounting function used
  here refers to the usual formulation in the field of reinforcement learning,
  although psychological experiments may also reveal other discounting regimes
  \citep{green2004discounting}.}
measures the cumulative reward of
following an action policy $\pi$.
%
Note that value buffering may be realized in the vmPFC by virtue of
this region's connections to the NAc of the reward system
~\citep{carlezon2009,haber1995,croxson2005quantitative}.

@@

\subsubsection{Optimal control via stochastic gradient descent}
Efficient learning of the entire set of model parameters can effectively be achieved
via stochastic \textit{gradient descent}, a universal algorithm for finding
local minima based on the first derivative of the optimization objective.
Stochastic here means that the gradient is estimated from batches of training samples,
which here corresponds to blocks of experience from the replay memory:

@@

\subsection{Summary and empirical hypotheses}

@@

The current account of DMN has multiple empirical consequences and predictions. Perhaps one of the first experimental venue concerns the neural correlates of the Bellman equation in the DMN. There are already relationship between the decomposition of consecutive action choices by the Bellman equation and neuroscientific insights:
specific neural activity in the dorsal prefrontal cortex (BA9)
was for instance linked to processing ``goal-tree sequences''
in human brain-imaging experiments
\citep{koechlin1999role, koechlin2000dissociating}.
Sub-goal exploration may require multi-task switching
between cognitive processes
as later parts of a solution
frequently depend on respective earlier steps in a given solution path, which
necessitates storage of expected intermediate outcomes.
As such,
``cognitive branching'' operations for nested processing of behavioral strategies
are likely to
entail secondary reallocation of attention and working-memory resources.
Further brain-imaging experiments corroborated
the prefrontal DMN to subserve
``processes related to the management and monitoring of sub-goals while
maintaining information in working memory''
\citep{braver2002role}
and to functionally couple with the hippocampus conditioned by ``deep versus shallow planning''
\citep{kaplan2017neural}.
%
Moreover,
neurological patients with lesions in this DMN region were reported
to be impaired in aspects of realizing ``multiple sub-goal scheduling''
\citep{burgess2000cognitive}.
Hence,
the various advanced human abilities subserved by the DMN, such as
planning and abstract reasoning, can be viewed to involve some form of
action-decision branching to enable higher-order executive control.
Here, we hypothesize a functional dissociation between computations pertaining to action policy versus adapting stimulus-value associations as we expect implementation in different subsystems of the DMN. First, we expect that fMRI signals in the right temporo-parietal junction relate to behavioral changes subsequent to adaptation in the action choice tendencies (policy matrix) involved in non-value-related prediction error. Second, fMRI signals in the ventromedial prefrontal cortex should relate to behavioral changes following adaptation in value estimation (value matrix) due to reward-related stimulus-value association. We finally expect that fMRI signals in the posteromedial cortex, as a potential global information integrator, are related to shifts in overt behavior based on previous adaptations in both policy or value estimation.

Beyond the Bellman equation and its potential underlying neural mechanisms, our process model of the DMN has also implication for experiments in neuroeconomy, especially for temporal discounting and continuous learning paradigms. Regarding the later, we for instance hypothesize a functional relationship between the DMN closely associated with the occurrence of stimulus-independent thoughts and the reward circuitry. During an iterative neuroeconomic two-player game, fMRI signals in the DMN could be used to predict reward-related signals in the nucleus accumbens across trials in a continuous learning paradigm. We expect that the more DMN activity is measured to be increased, supposedly the higher the tendency for stimulus-independent thoughts, the more the fMRI signals in the reward circuits should be independent of the reward context in the current sensory environment. In the case of temporal discounting, we hypothesize that the relevant time horizon is modulated by various factors such as age, acute stress and time-enduring impulsivity traits
\citep{luksys2009stress, haushofer2014psychology}.
Using a temporal discounting experiment, it can be quantified how the time horizon is affected at the behavioral level and then traced back to its corresponding neural representation. Such experimental investigation can be designed to examine between-group and within-group effects (e.g., impulsive population like chronic gamblers or drug addicts); and brought in context with the participant’s age and personality traits.
% To exam-ine this question, Fellows and Farah (2005) investigated the effects of dorsolateral frontal (DLF) and ventromedial frontal (VMF) lobe damage on two aspects of temporal foresight—namely, temporal discounting and the futuretime perspective (measuredby an adaptation of Wallace’s [1956] list of future life events).

@@

Another prediction from the current model is that the HC may contribute to generating
perturbed action-transition-state-reward samples
as batches of pseudo-experience
(i.e., recalled, hypothesized, and forecasted scenarios).
The small variations in these experience samplings allow searching
a larger space of model parameters and possible experiences.
Taken to its extreme, stochastic recombination of experience
building blocks can further optimize the behavior of the RL agent
by learning from scenarios in the environment that the agent might
encounter only very rarely or never.
An explanation is thus offered for experiencing seemingly familiar situations that
a human has however never actually encountered (i.e., d\'{e}j\`{a} vu effect).
While such a situation may not have been experienced in the physical world,
the DMN may have previously stochastically generated, evaluated, and adapted to
such a randomly synthesized event.
Generated representations arguably are ``internally manipulable,
and can be used for attempting actions internally,
before or instead of acting in the external reality,
and in diverse goal and sensory contexts, i.e. even outside the context in which they were learned''
\citep{pezzulo2011grounding}.
In the context of scarce environmental input and feedback
(e.g., mind-wandering or sleep),
mental scene construction allows pseudo-experiencing possible
future scenarios and action outcomes.

This interplay between memory and “mind-searching” also supports that experience replay for browsing problem solutions subserved by the DMN is contributing to choice behavior in mice. Hippocampal single-cell recordings have shown that neural patterns during experimental choice behavior are reiterated during sleep and before making analogous choices in the future. We hypothesize that, in addition to the hippocampus, there is a necessity of cortical DMN regions for “mind-searching” candidate actions during choice behavior. It can be experimentally corroborated by causal disruption of DMN regions, such as by circumscribed brain lesion or optogenetic intervention in the inferior parietal and prefrontal cortices.
From the perspective of a RL agent,
prediction in the DMN reduces to
generalization of
policy and value computations from sampled experiences to
successful action choices and reward predictions in future states.
As such,
plasticity in the DMN arises naturally.
If an agent behaving optimally in a certain environment moves
to new, yet unexperienced environment, reward prediction errors will
largely increase.
This feedback will lead to adaptation of policy considerations
and value estimations until the intelligent system converges to a
new steady state of optimal action decisions in a volatile world.

A last empirical perspective concerns the role of synaptic epigenesis in the creation of new entry in policy matrix. Indeed, we did not adress here the additional layer of learning which concerns the addition of new entries in the state and action spaces. Extension of the action repertoire could be biologically realized by synaptic epigenesis \citep{gisiger_acquisition_2005}. Indeed, the tuning of synaptic weights through learning can stabilize additional patterns of activity by creating new attractors in the neural dynamics landscape \citep{takeuchi_synaptic_2014}. Those attractors can then constrain both the number of factors taken into account by decision processes and the possible behaviors of the agent \citep{wang_decision_2008}. To examine this potential higher-level mechanism, we propose to probe how synaptic epigenesis is related to neural correlates underlying policy matrix updates: in humans the changes of functional connectivity between DMN regions can be investigated following a temporal discounting experiment and in monkeys or rodents anterograde tracing can be used to study how homolog regions of the DMN present increased synaptic changes compare to other parts of the brain.

@@

\begin{figure}[!h]
  \includegraphics[width=1\linewidth]{VennDiagram-2018-08-11.pdf}
  \caption{\textbf{Situating Markov Decision Processes among other accounts of default mode function.}
  The Venn diagram summarizes the relationship between four previously proposed explanations for the functional role of the DMN and our present account. Viewing empirical findings in the DMN from the MDP viewpoint incorporates important aspects of the free energy principle, predictive coding, sentinal hypothesis, and semantic hypothesis.
  The MDP account may reconcile several strengths of these functional accounts in a process model that simultaneously acknowledges environmental input and behavioral choices as well as the computational and algorithmic properties (How? and What?) underlying higher-order control of the organism.}
  \label{fig:VennDiagram}
\end{figure}
